---
title: Stat 502 Project Proposal
date: November 2, 2016
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = T)
options(digits = 3)
df <- log(data.frame(t(apply(read.table('proposal-results.txt', header = T), 1, function(x) x / x[1]))))
```

## Background

Python is one of the most popular programming languages, with many applications in scientific computing.
Thus, maximizing the performance of Python code is of particular interest.
Because Python is an interpreted language, its performance depends on the particular runtime used.
Interpreted languages often rely on a just-in-time (JIT) compiler to dynamically optimize the code at runtime, which the official CPython interpreter lacks.
Fortunately, there are many options for Python runtimes besides the official CPython implementations for versions 2 and 3, including PyPy, Jython, and IronPython.
PyPy uses its own JIT compiler, while Jython and IronPython are built on top of the Java and .NET frameworks, respectively, which encompass their own JIT compilers.
In this study we hope to determine whether the use of a runtime with a JIT compiler offers a performance benefit over the official CPython implementations.

## Pilot study



```{r, dev='pdf', fig.cap=''}
boxplot(df)
```

```{r}
anova(lm(values ~ ind, data=stack(df)))
```

## Proposal for larger experiment



### Power calculation

